---
title: "A Short History of Artificial Intelligence"
subtitle: "From Neurons to Transformers"
author: "Matteo Lisi"
format:
  revealjs:
    logo: ../img/logo-small-london-cmyk.jpg
    footer: "PS3192 25-26"
    incremental: true  
    auto-stretch: false
    code-fold: false   # don’t fold
    code-line-numbers: false
    theme: [default, ../matteo_rhul.css]
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(ggplot2)
library(igraph)
library(ggraph)
library(tidyverse)
```

## Overview {.center}

> *"Modern AI is built on neural networks — loosely inspired by the brain, but increasingly diverging from it."*

This session traces the key milestones from the first artificial neuron (1943) to the Transformer (2017), asking at each step: **how brain-like is this?**

---

## The Artificial Neuron — McCulloch & Pitts (1943) {.smaller}

:::: {.columns}

::: {.column width="45%"}
**The idea:**

- A simple unit of computation inspired by the biological neuron
- Takes **binary inputs**, computes a **weighted sum**, fires if it exceeds a threshold
- Can be combined to implement **any Boolean logic**

**But:** weights are fixed — it cannot learn
:::

::: {.column width="55%"}
```{r mcp-neuron, fig.height=4, fig.width=5}
par(bg = "transparent", mar = c(1, 1, 1.5, 1))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 10))

# Input nodes
input_x <- 1.2; input_ys <- c(7.5, 5, 2.5)
node_col <- "#56B4E9"; neuron_col <- "#E69F00"

for (y in input_ys) {
  symbols(input_x, y, circles = 0.5, add = TRUE, inches = FALSE,
          bg = node_col, fg = "white", lwd = 2)
}
text(input_x, 9.2, "Inputs  x₁, x₂, x₃", col = "white", cex = 0.85, font = 2)

# Weights on arrows
w_labels <- c("w₁", "w₂", "w₃")
neuron_x <- 5.5; neuron_y <- 5
for (i in seq_along(input_ys)) {
  arrows(input_x + 0.5, input_ys[i], neuron_x - 0.6, neuron_y,
         length = 0.1, col = "white", lwd = 1.5)
  mx <- (input_x + 0.5 + neuron_x - 0.6) / 2
  my <- (input_ys[i] + neuron_y) / 2
  text(mx, my + 0.4, w_labels[i], col = "#F0E442", cex = 0.8, font = 3)
}

# Neuron body
symbols(neuron_x, neuron_y, circles = 0.7, add = TRUE, inches = FALSE,
        bg = neuron_col, fg = "white", lwd = 2)
text(neuron_x, neuron_y, "Σ ≥ θ?", col = "white", cex = 0.75, font = 2)
text(neuron_x, 9.2, "Neuron", col = "white", cex = 0.85, font = 2)

# Output
output_x <- 8.5
arrows(neuron_x + 0.7, neuron_y, output_x - 0.3, neuron_y,
       length = 0.12, col = "white", lwd = 2)
symbols(output_x, neuron_y, circles = 0.4, add = TRUE, inches = FALSE,
        bg = "#009E73", fg = "white", lwd = 2)
text(output_x, neuron_y, "0/1", col = "white", cex = 0.7, font = 2)
text(output_x, 9.2, "Output", col = "white", cex = 0.85, font = 2)

title("The McCulloch–Pitts Neuron (1943)", col.main = "white", cex.main = 1)
```
:::

::::

---

## Hebbian Learning (1949) & The Perceptron (1962) {.smaller}

:::: {.columns}

::: {.column width="48%"}
**Donald Hebb (1949)**

> *"Neurons that fire together, wire together"*

- Learning as **synaptic strengthening** between co-active neurons
- A biological principle that became a foundation for computational learning

**Frank Rosenblatt (1962)**

- The **Perceptron**: a single layer network that can **learn weights from data**
- Weights encode knowledge — a profound shift
- The first algorithm to learn a linear decision boundary
:::

::: {.column width="52%"}
```{r perceptron, fig.height=4.5, fig.width=5}
par(bg = "transparent", mar = c(2, 1, 2, 1))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 10))

input_xs <- rep(1.5, 4)
input_ys <- c(8, 5.5, 3.5, 1)
out_x <- 8; out_y <- 5

node_col <- "#56B4E9"
out_col  <- "#E69F00"

# Inputs
for (y in input_ys) {
  symbols(1.5, y, circles = 0.55, add = TRUE, inches = FALSE,
          bg = node_col, fg = "white", lwd = 2)
}
text(1.5, 9.4, "Inputs", col = "white", cex = 0.85, font = 2)

# Arrows with weight labels
ws <- c("w₁", "w₂", "w₃", "w₄")
for (i in seq_along(input_ys)) {
  arrows(2.05, input_ys[i], out_x - 0.55, out_y,
         length = 0.09, col = "#cccccc", lwd = 1.4)
  mx <- (2.05 + out_x - 0.55) / 2
  my <- (input_ys[i] + out_y) / 2
  text(mx + 0.2, my + 0.35, ws[i], col = "#F0E442", cex = 0.72, font = 3)
}

# Output node
symbols(out_x, out_y, circles = 0.7, add = TRUE, inches = FALSE,
        bg = out_col, fg = "white", lwd = 2)
text(out_x, out_y + 0.05, "ŷ", col = "white", cex = 1.1, font = 2)
text(out_x, 9.4, "Output", col = "white", cex = 0.85, font = 2)

# Learning arrow feedback
arrows(out_x, out_y - 0.75, out_x - 2, out_y - 2.5,
       length = 0.1, col = "#CC79A7", lwd = 1.8, lty = 2)
text(out_x - 1.2, out_y - 2.0, "error → update w", col = "#CC79A7", cex = 0.72, font = 3)

# Input labels
xlabs <- c("age", "BMI", "fever", "CT scan")
for (i in seq_along(input_ys)) {
  text(0.1, input_ys[i], xlabs[i], col = "#aaaaaa", cex = 0.65, adj = 0)
}

title("Perceptron — learns weights from data", col.main = "white", cex.main = 1)
```
:::

::::

---

## Patient Vectors — A Concrete Example {.smaller}

*Inspired by Anil Ananthaswamy, "I, Human" / "Why Machines Learn"*

Each patient is represented as a **vector in 6-dimensional space**:

| Feature | Value |
|---|---|
| Age | continuous |
| Body mass | continuous |
| Breathing difficulty | 0 / 1 |
| Fever | 0 / 1 |
| Diabetes | 0 / 1 |
| Chest CT scan score | continuous |

**Target (label):** Does the patient improve after 3 days — or remain on ventilator support?

::: {.callout-tip}
## Why this matters
The perceptron finds a **hyperplane** in this 6D space that separates "improves" from "remains sick." The **weights are the knowledge** — learned automatically from labelled examples.
:::

---

## Limits — The XOR Problem & Minsky–Papert (1969) {.smaller}

:::: {.columns}

::: {.column width="50%"}
**The good news:**

*Perceptron Convergence Theorem* — if a linear solution exists, the perceptron **is guaranteed to find it**.

**The bad news:**

Many real problems are **not linearly separable** — the classic case is XOR:

| x₁ | x₂ | XOR |
|---|---|---|
| 0 | 0 | 0 |
| 0 | 1 | 1 |
| 1 | 0 | 1 |
| 1 | 1 | 0 |

No single straight line can separate the 1s from the 0s.

Minsky & Papert (1969) formalised these limitations — **setting off the first AI Winter**.
:::

::: {.column width="50%"}
```{r xor, fig.height=4, fig.width=4.5}
par(bg = "transparent", mar = c(3, 3, 2, 1))
plot(c(0,1,0,1), c(0,1,1,0),
     pch = c(1, 1, 16, 16), cex = 3.5,
     col = c("#56B4E9","#E69F00","#E69F00","#56B4E9"),
     xlim = c(-0.3, 1.3), ylim = c(-0.3, 1.3),
     xlab = "x₁", ylab = "x₂",
     main = "XOR — not linearly separable",
     col.main = "white", col.lab = "white", col.axis = "white",
     bg = "transparent", axes = TRUE)
grid(col = "#444444")
legend("topright",
       legend = c("Output = 0", "Output = 1"),
       pch = c(1, 16),
       col = c("#56B4E9", "#E69F00"),
       text.col = "white", bg = "transparent", box.col = "transparent")
axis(1, col.axis = "white", col = "white")
axis(2, col.axis = "white", col = "white")
box(col = "white")
```
:::

::::

---

## The First AI Winter: 1974–1980 {.center}

::: {.callout-warning}
## The classic winter pattern
Not "no progress" — but **expectations rising much faster than capabilities**.
:::

In the 1970s, AI energy shifted toward **symbolic AI & Expert Systems**:

- Knowledge represented explicitly as **rules, logical formulas, frames, semantic networks**
- Systems like MYCIN (medical diagnosis) showed impressive narrow performance
- But: **brittle, hard to maintain, couldn't generalise**

Buyers expected adaptability and robustness — they got carefully hand-coded rules.

::: {.fragment}
> A second AI Winter followed in the **late 1980s** for the same reason: hype outran capability.
:::

---

## Backpropagation — Rumelhart, Hinton & Williams (1982) {.smaller}

:::: {.columns}

::: {.column width="50%"}
The key insight that made **multi-layer networks trainable**:

- Compute how much each weight contributed to the error
- Propagate that gradient **backwards** through the layers
- Update weights using gradient descent

This allowed networks to learn **internal representations** — not just input-output mappings.

It also implicitly solved XOR: add a hidden layer, and you can represent non-linear boundaries.
:::

::: {.column width="50%"}
```{r feedforward, fig.height=5, fig.width=4.8}
par(bg = "transparent", mar = c(1, 1, 2, 1))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 10))

layers <- list(
  input  = list(x = 1.5, ys = c(8, 6, 4, 2),  col = "#56B4E9", label = "Input"),
  hidden = list(x = 5,   ys = c(7.5, 5.5, 3.5),col = "#E69F00", label = "Hidden"),
  output = list(x = 8.5, ys = c(6.5, 4, 1.5),  col = "#009E73", label = "Output")
)

# Forward connections
for (i in seq_along(layers$input$ys)) {
  for (j in seq_along(layers$hidden$ys)) {
    lines(c(layers$input$x + 0.42, layers$hidden$x - 0.42),
          c(layers$input$ys[i],    layers$hidden$ys[j]),
          col = "#888888", lwd = 0.9)
  }
}
for (i in seq_along(layers$hidden$ys)) {
  for (j in seq_along(layers$output$ys)) {
    lines(c(layers$hidden$x + 0.42, layers$output$x - 0.42),
          c(layers$hidden$ys[i],    layers$output$ys[j]),
          col = "#888888", lwd = 0.9)
  }
}

# Nodes
for (nm in names(layers)) {
  l <- layers[[nm]]
  for (y in l$ys) {
    symbols(l$x, y, circles = 0.42, add = TRUE, inches = FALSE,
            bg = l$col, fg = "white", lwd = 1.8)
  }
  text(l$x, 9.5, l$label, col = "white", cex = 0.85, font = 2)
}

# Backprop arrow
arrows(7.8, 1.0, 2.2, 1.0, length = 0.12,
       col = "#CC79A7", lwd = 2, lty = 2)
text(5, 0.4, "← backpropagation (error gradient)", col = "#CC79A7", cex = 0.72, font = 3)

title("Feedforward Network + Backprop", col.main = "white", cex.main = 1)
```
:::

::::

---

## Convolutional Neural Networks — LeCun et al. (1989) {.smaller}

:::: {.columns}

::: {.column width="50%"}
**Brain-inspired architecture for vision:**

- **Receptive fields**: each unit responds to a local patch, mimicking visual cortex
- **Weight sharing**: the same filter scans the whole image → far fewer parameters
- **Pooling**: spatial downsampling, building invariance
- **Hierarchy**: edges → textures → parts → objects — "features of features"

Biologically motivated, practically transformative — but the hierarchy is steeper and more explicit than in biological vision.
:::

::: {.column width="50%"}
```{r cnn-diagram, fig.height=5, fig.width=5}
par(bg = "transparent", mar = c(1, 1, 2, 1))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 10))

# Layer blocks
blocks <- data.frame(
  x    = c(0.5, 2.8, 5.1, 7.4),
  w    = c(1.8, 1.8, 1.5, 1.6),
  h    = c(4.0, 3.0, 2.0, 1.2),
  col  = c("#56B4E9","#E69F00","#009E73","#CC79A7"),
  lab  = c("Input\nimage", "Conv\nlayer", "Pooling\n+ Conv", "Fully\nconnected"),
  stringsAsFactors = FALSE
)

cy <- 5  # centre y
for (i in 1:nrow(blocks)) {
  b <- blocks[i, ]
  rect(b$x, cy - b$h/2, b$x + b$w, cy + b$h/2,
       col = b$col, border = "white", lwd = 1.5)
  text(b$x + b$w/2, cy, b$lab, col = "white", cex = 0.72, font = 2)
  text(b$x + b$w/2, cy + b$h/2 + 0.4, c("28×28","24×24","12×12","")[i],
       col = "#aaaaaa", cex = 0.62)
}

# Arrows between blocks
for (i in 1:(nrow(blocks)-1)) {
  b1 <- blocks[i, ]; b2 <- blocks[i+1, ]
  arrows(b1$x + b1$w + 0.05, cy, b2$x - 0.05, cy,
         length = 0.1, col = "white", lwd = 1.8)
}

# Labels under
feature_labels <- c("pixels", "edges &\norientations", "textures &\nparts", "class\nscores")
for (i in 1:nrow(blocks)) {
  b <- blocks[i, ]
  text(b$x + b$w/2, cy - b$h/2 - 0.55, feature_labels[i],
       col = "#cccccc", cex = 0.62)
}

title("Convolutional Network — Hierarchical Features", col.main = "white", cex.main = 1)
```
:::

::::

---

## Recurrent Networks & LSTM (1997) {.smaller}

:::: {.columns}

::: {.column width="48%"}
**The problem:** sequences have **temporal dependencies**

- Language, speech, time series — earlier inputs matter later
- Standard feedforward nets have no memory

**Recurrent networks:** hidden state is fed back as input at the next step

**LSTM (Hochreiter & Schmidhuber, 1997):**

- Gated memory cells allow gradients to flow over **long distances** in time
- Solved the "vanishing gradient" problem
- Brain-inspired: recurrence echoes working memory

*Still: biological recurrence is richer, messier, and multi-scale*
:::

::: {.column width="52%"}
```{r rnn-diagram, fig.height=4.5, fig.width=5}
par(bg = "transparent", mar = c(2, 1, 2, 1))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 10))

# Three unrolled RNN steps
xs <- c(2, 5, 8); hy <- 6; iy <- 2.5; oy <- 9

node_cols <- c("#56B4E9","#E69F00","#009E73")
time_labs <- c("t-1", "t", "t+1")

for (i in seq_along(xs)) {
  # hidden node
  symbols(xs[i], hy, circles = 0.6, add = TRUE, inches = FALSE,
          bg = node_cols[i], fg = "white", lwd = 2)
  text(xs[i], hy, "h", col = "white", cex = 0.9, font = 2)

  # input node
  symbols(xs[i], iy, circles = 0.45, add = TRUE, inches = FALSE,
          bg = "#555555", fg = "white", lwd = 1.5)
  text(xs[i], iy, paste0("x\n",time_labs[i]), col = "white", cex = 0.6)

  # output node
  symbols(xs[i], oy, circles = 0.45, add = TRUE, inches = FALSE,
          bg = "#9E5FB8", fg = "white", lwd = 1.5)
  text(xs[i], oy, paste0("y\n",time_labs[i]), col = "white", cex = 0.6)

  # input → hidden
  arrows(xs[i], iy + 0.45, xs[i], hy - 0.6, length = 0.1, col = "white", lwd = 1.5)
  # hidden → output
  arrows(xs[i], hy + 0.6, xs[i], oy - 0.45, length = 0.1, col = "white", lwd = 1.5)
}

# Recurrent connections between hidden states
for (i in 1:(length(xs)-1)) {
  arrows(xs[i] + 0.6, hy, xs[i+1] - 0.6, hy,
         length = 0.1, col = "#F0E442", lwd = 2)
}
text(5, 7.3, "recurrent connection →", col = "#F0E442", cex = 0.72, font = 3)

# Self-loop on middle node
theta <- seq(pi*0.2, pi*1.8, length.out = 50)
r <- 0.9
loop_x <- xs[2] + r * cos(theta)
loop_y <- hy + r * 0.5 + 0.3 * sin(theta)
lines(loop_x, loop_y, col = "#F0E442", lwd = 1.5, lty = 2)

title("Recurrent Network — Sequence Memory", col.main = "white", cex.main = 1)
```
:::

::::

---

## 2012 — The Deep Learning Revolution {.center}

**AlexNet wins ImageNet** by more than **10 percentage points** over all competitors.

::: {.incremental}
- Trained on GPUs — showing that hardware mattered as much as algorithms
- Used **ReLU activations** and **dropout regularisation** — engineering over biology
- Triggered enormous investment from industry (Google, Microsoft, Meta, …)
- The moment neural networks moved from academic curiosity to industrial infrastructure
:::

::: {.fragment}
> *"The data is there. The compute is there. Now the architecture works."*
:::

---

## 2013 — Word2Vec: Meaning as Geometry {.smaller}

:::: {.columns}

::: {.column width="55%"}
**Mikolov et al. (2013):** learn dense word vectors from large text corpora

- Words with similar meaning → **nearby vectors** in high-dimensional space
- Semantic relationships encoded as **directions**:
  - King − Man + Woman ≈ Queen
  - Paris − France + Italy ≈ Rome

**Why this matters:**

- Language is no longer a bag of discrete atoms
- Meaning is **distributed** across many dimensions — closer to how neuroscience thinks about distributed coding
- Downstream tasks (classification, translation) suddenly became much easier
:::

::: {.column width="45%"}
```{r word2vec, fig.height=4.5, fig.width=4.5}
par(bg = "transparent", mar = c(3, 3, 2, 1))
words <- list(
  King   = c(0.1, 0.9), Man  = c(0.05, 0.4),
  Queen  = c(0.75, 0.85), Woman = c(0.7, 0.35),
  France = c(0.15, -0.2), Paris = c(0.25, 0.15),
  Italy  = c(0.7, -0.25), Rome  = c(0.8, 0.1)
)
df <- do.call(rbind, words) |> as.data.frame()
colnames(df) <- c("x","y")
df$word <- rownames(df)

plot(df$x, df$y, type = "n",
     xlim = c(-0.1, 1.0), ylim = c(-0.5, 1.1),
     xlab = "Dimension 1 (gender?)", ylab = "Dimension 2 (royalty?)",
     col.lab = "white", col.axis = "white", main = "Word Vectors (schematic)",
     col.main = "white", axes = TRUE)
grid(col = "#444444")
box(col = "white")

# Analogy arrows
arrows(df["Man","x"], df["Man","y"], df["King","x"], df["King","y"],
       col="#E69F00", lwd=1.4, length=0.1, lty=2)
arrows(df["Woman","x"], df["Woman","y"], df["Queen","x"], df["Queen","y"],
       col="#E69F00", lwd=1.4, length=0.1, lty=2)
arrows(df["France","x"], df["France","y"], df["Paris","x"], df["Paris","y"],
       col="#56B4E9", lwd=1.4, length=0.1, lty=2)
arrows(df["Italy","x"], df["Italy","y"], df["Rome","x"], df["Rome","y"],
       col="#56B4E9", lwd=1.4, length=0.1, lty=2)

pch_vals <- c(King=17,Man=1,Queen=17,Woman=1,
               France=15,Paris=15,Italy=15,Rome=15)
col_vals <- c(King="#E69F00",Man="#E69F00",Queen="#E69F00",Woman="#E69F00",
               France="#56B4E9",Paris="#56B4E9",Italy="#56B4E9",Rome="#56B4E9")
for (w in rownames(df)) {
  points(df[w,"x"], df[w,"y"], pch = pch_vals[w], cex = 1.8,
         col = col_vals[w])
  text(df[w,"x"], df[w,"y"] + 0.09, w, col = "white", cex = 0.7)
}
axis(1, col.axis="white", col="white")
axis(2, col.axis="white", col="white")
```
:::

::::

---

## 2017 — "Attention Is All You Need" {.smaller}

The **Transformer** architecture (Vaswani et al.) marks a decisive break.

It **dispensed entirely** with the two most brain-like prior principles:

::: {.columns}

::: {.column width="50%"}
**What it dropped:**

❌ **Recurrence** — no sequential hidden state  
❌ **Convolutions** — no local receptive fields

**What makes it un-brain-like:**

- **Global pairwise attention:** every token interacts with every other token simultaneously
- **Multi-head attention:** no known mapping to neural circuitry
- **Positional encoding:** brain does not add sinusoidal tags to inputs to track order
:::

::: {.column width="50%"}
**Why it works:**

✅ Massively parallelisable on GPUs  
✅ Scales with data and compute  
✅ No vanishing gradient through time  
✅ Long-range dependencies handled directly

> *Performance improves precisely by abandoning biological constraints.*
:::

::::

---

## The Trajectory: From Biology to Engineering {.smaller}

```{r timeline, fig.height=4.5, fig.width=10}
par(bg = "transparent", mar = c(4, 1, 2, 1))
plot.new()
plot.window(xlim = c(1940, 2022), ylim = c(0, 4))

# Timeline axis
lines(c(1943, 2018), c(1.5, 1.5), col = "white", lwd = 2)
arrows(2018, 1.5, 2021, 1.5, length = 0.1, col = "white", lwd = 2)

events <- data.frame(
  year  = c(1943, 1949, 1962, 1969, 1974, 1982, 1989, 1997, 2012, 2013, 2017),
  label = c("MCP\nNeuron", "Hebb\nLearning", "Perceptron", "Minsky\n& Papert",
            "1st AI\nWinter", "Backprop", "CNN\nLeCun", "LSTM",
            "AlexNet", "Word2Vec", "Transformer"),
  above = c(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE, TRUE, FALSE, TRUE, FALSE, TRUE),
  bio   = c(1, 1, 1, 1, 0, 1, 1, 1, 0.5, 0.5, 0),
  stringsAsFactors = FALSE
)

bio_cols <- colorRampPalette(c("#CC79A7","#E69F00","#56B4E9"))(3)

for (i in 1:nrow(events)) {
  e <- events[i,]
  col_i <- if (e$bio == 1) "#56B4E9" else if (e$bio == 0.5) "#E69F00" else "#CC79A7"
  
  points(e$year, 1.5, pch = 21, cex = 2.2, bg = col_i, col = "white", lwd = 1.5)
  
  if (e$above) {
    segments(e$year, 1.5, e$year, 2.8, col = "#777777", lwd = 1)
    text(e$year, 3.0, e$label, col = "white", cex = 0.58, font = 2, adj = 0.5)
    text(e$year, 1.1, e$year, col = "#aaaaaa", cex = 0.55)
  } else {
    segments(e$year, 1.5, e$year, 0.3, col = "#777777", lwd = 1)
    text(e$year, 0.1, e$label, col = "white", cex = 0.58, font = 2, adj = 0.5)
    text(e$year, 1.9, e$year, col = "#aaaaaa", cex = 0.55)
  }
}

# Legend
legend(1943, 4.0, legend = c("Brain-inspired", "Hybrid", "Engineering-first"),
       pch = 21, pt.bg = c("#56B4E9","#E69F00","#CC79A7"),
       col = "white", text.col = "white", bg = "transparent",
       box.col = "transparent", horiz = TRUE, pt.cex = 1.5, cex = 0.75)

title("Key Milestones in Neural Network AI", col.main = "white", cex.main = 1.1)
```

---

## The Big Picture {.center}

::: {.callout-note}
## A recurring pattern
Early deep learning borrowed broad principles from biology — distributed representation, learned connectivity, hierarchy, recurrence — while progressively **replacing biological realism with engineering abstractions**.
:::

::: {.fragment}
The Transformer marks the moment when that trend becomes **explicit**:

> *Performance improves precisely by abandoning the most brain-like architectural constraints.*
:::

::: {.fragment}
**So: how brain-like is modern AI?**

Less so with every generation — and more capable for it.
:::

---

## Key Takeaways {.smaller}

| Milestone | Year | Brain-like? |
|---|---|---|
| McCulloch–Pitts neuron | 1943 | ✅ Inspired by neuroscience |
| Perceptron | 1962 | ✅ Learns like Hebb's rule |
| Backpropagation | 1982 | ⚠️ Mathematically elegant, biologically implausible |
| CNN | 1989 | ✅ Mimics visual cortex hierarchy |
| LSTM | 1997 | ⚠️ Captures recurrence, but gates are not biological |
| AlexNet | 2012 | ⚠️ Deep hierarchy, but GPU-optimised |
| Word2Vec | 2013 | ⚠️ Distributed coding — sort of |
| Transformer | 2017 | ❌ Deliberately non-biological |

::: {.fragment}
*The field started by asking "how does the brain compute?" — and ended up asking "what computes best?"*
:::
