---
title: "Artificial intelligence"
subtitle: "<br>PS3192: Data science for psychology"
author: "Matteo Lisi"
format:
  revealjs:
    logo: ../img/logo-small-london-cmyk.jpg
    footer: "PS3192 25-26"
    incremental: true  
    auto-stretch: false
    code-fold: false   # don’t fold
    code-line-numbers: false
    theme: [default, ../matteo_rhul.css]
editor: source
keep-md: true
filters: [../bg_style.lua]
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggdag)
library(dagitty)
```

## What is modern AI?

- Most modern AI is based on **artificial neural networks**
- These are systems made of many simple computational units connected together
- They are **loosely inspired by the brain**, but only in a broad sense
- The key idea is simple:
  - inputs are represented as numbers
  - the network transforms them through layers of computation
  - the connection strengths (**weights**) are learned from data

::: notes
Emphasise that “brain-inspired” does not mean “brain-realistic”.
The inspiration is historical and conceptual, not a faithful biological model.
:::

## How brain-like is it?

- Early neural networks borrowed some broad ideas from neuroscience:
  - neurons as simple units
  - learning by changing connections
  - hierarchical processing
  - recurrent feedback
- But modern AI is mostly driven by **engineering choices that scale well**
- Over time, performance improved by becoming **less biologically realistic**

> A useful summary: neural networks began as rough brain-inspired models, but modern AI is better understood as a family of powerful statistical function approximators.

## A very simple artificial neuron

- An artificial neuron:
  - receives several inputs
  - multiplies each input by a weight
  - adds them together
  - applies an activation rule
- In compact form:

$$
z = w_1x_1 + w_2x_2 + \dots + w_px_p + b
$$

$$
y = f(z)
$$

- The weights determine how important each input is
- Learning means adjusting those weights from data

## Artificial neuron: diagram

```{r, echo=FALSE, fig.width=9, fig.height=4}
par(mar = c(0,0,0,0))
plot(0, 0, type = "n", xlim = c(0, 10), ylim = c(0, 6),
     axes = FALSE, xlab = "", ylab = "")

# inputs
points(rep(1.5, 4), c(1.2, 2.4, 3.6, 4.8), pch = 19, cex = 1.4)
text(1, c(1.2, 2.4, 3.6, 4.8), labels = c("x1", "x2", "x3", "x4"), cex = 1.1)

# neuron
symbols(5, 3, circles = 0.8, inches = FALSE, add = TRUE, bg = "grey95")
text(5, 3, "Σ\n+ f()", cex = 1)

# output
points(8.5, 3, pch = 19, cex = 1.4)
text(9, 3, "y", pos = 4, cex = 1.1)

# arrows
arrows(1.8, 1.2, 4.2, 2.6, length = 0.08)
arrows(1.8, 2.4, 4.2, 2.85, length = 0.08)
arrows(1.8, 3.6, 4.2, 3.15, length = 0.08)
arrows(1.8, 4.8, 4.2, 3.4, length = 0.08)
arrows(5.8, 3, 8, 3, length = 0.08)

# weights
text(3, c(1.8, 2.6, 3.4, 4.2), labels = c("w1", "w2", "w3", "w4"), cex = 0.9)

# bias
arrows(5, 5.5, 5, 3.9, length = 0.08)
text(5.25, 5.5, "bias b", pos = 4, cex = 0.9)